---
title: "CONW trend models"
format: html
editor: visual
---

## Collection of trend models applied to CONW

Connecticut Warbler population trends are partly dependent on the model and the underlying geographic stratifications.

To explore the among-model variation in our understanding of the species' population trends, we fit the data to two models, each at two different stratifications to estimate trends at different spatial-grains. The two models are a GAMYE and a first-difference model (). The two stratifications are: 1) coarse-grained stratification based on the intersection of political jurisdictions (states, provinces, territories) and Bird Conservation Regions, (BCRs) that we refer to here as the `"bbs_usgs"` and a finer-grained stratification based on a latitude by longitude 1-degree grid-cell that we refer to here as `"latlong"`.

## Model fitting

We used the R package `bbsBayes2`.

```{r, message=FALSE}
library(bbsBayes2)
library(patchwork)
library(tidyverse)
library(sf)
source("functions/trend_colour_palette_apply.r")

models <- c("gamye","first_diff")
stratifications <- c("latlong","bbs_usgs")
species <- "Connecticut Warbler"
refit <- FALSE
if(refit){
for(mod in models){
  for(stratification in stratifications){
    min_n <- ifelse(stratification == "latlong",1,3)
    s <- stratify(by = stratification,
                  species = species)
    p <- prepare_data(s,
                      min_n_routes = min_n,
                      min_max_route_years = 2)
    ps <- prepare_spatial(p,strata_map = load_map(stratification))
    
    pm <- prepare_model(ps,
                  model = mod,
                  model_variant = "spatial")
    
    saveRDS(pm,paste0("data/",mod,"_",stratification,".rds"))
    
    m <- run_model(pm,
                   output_basename = paste0(mod,"_",stratification),
                   output_dir = "output",
                   iter_sampling = 3000)
    
    summ <- get_summary(m) %>% 
      mutate(model = mod,
             stratification = stratification)
    saveRDS(summ,paste0("output/parameter_summary_",mod,"_",stratification,".rds"))
  
  }
}
}


```

We prepare the data setting the minimum number of routes to different values depending on the stratification used. For the coarse-grain stratification, we use the 3-route minimum cut-off to reflect the thresholds used by the CWS and USGS in their annual analyses. For the fine-grain stratification, we use a minimum of 1-route, to include the largest number of strata and to reflect the underlying design of the BBS. For all estimates, we also set a higher minimum for the number of non-zero observations on a given route to ensure that each included route and stratum has more reliable data for estimating change in abundance over time.

Then we estimate population trajectories and trends, comparing them across these two models and stratifications.

```{r}

fits <- vector(mode = "list",4)
names(fits) <- paste0(rep(models,each = 2),"_",rep(stratifications,2))
summ_out <- NULL
indices_out <- NULL
trends_out <- NULL


for(mod in models){
  for(stratification in stratifications){

    tmp <- readRDS(paste0("output/",mod,"_",stratification,".rds"))

    # summ <- get_summary(tmp) %>% 
    #   mutate(model = mod,
    #          stratification = stratification)
  
      summ <- readRDS(paste0("output/parameter_summary_",mod,"_",stratification,".rds"))
  summ_out <- bind_rows(summ_out,summ)
    fits[[paste0(mod,"_",stratification)]] <- tmp
  



## group latlong strata into original BBS strata
if(stratification == "latlong"){
  
    bbs_usgs <- load_map("bbs_usgs")%>%
      select(-area_sq_km) %>% 
      rename(bbs_strata = strata_name,
             Country = country)
    
    
    strata_join <- load_map(stratification) %>% 
    filter(strata_name %in% tmp$meta_strata$strata_name) %>%
    sf::st_join(.,bbs_usgs,
                largest = TRUE,
                left = TRUE) %>% # intersection
    sf::st_drop_geometry() %>%
    mutate(bbs_strata = ifelse(is.na(bbs_strata),"other",bbs_strata))

      # generate trajectories for original BBS strata
  inds_tmp <- generate_indices(tmp,alternate_n = "n",
                                    regions = c("Country","bbs_strata"),
                                    regions_index = strata_join,
                                    hpdi = TRUE,
                                    start_year = 1970)
    saveRDS(inds_tmp,paste0("output/inds_comp_",mod,"_",stratification,".rds"))
    
    indices_tmp <- inds_tmp$indices %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "composite",
             type = "full")
    indices_out <- bind_rows(indices_out,
                          indices_tmp)
    
    tt_long <- generate_trends(inds_tmp,
                               min_year = 1970,
                               prob_decrease = c(0,30,50))
    tt_short <- generate_trends(inds_tmp,
                               min_year = 2012,
                               prob_decrease = c(0,30,50))
        tt_pre_90 <- generate_trends(inds_tmp,
                               min_year = 1970,
                               max_year = 1990,
                               prob_decrease = c(0,30,50))
            tt_post_90 <- generate_trends(inds_tmp,
                               min_year = 1990,
                               prob_decrease = c(0,30,50))
    tt_t <- tt_long$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "composite",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
    
    tt_t <- tt_short$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "composite",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
    
        tt_t <- tt_pre_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "composite",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
        tt_t <- tt_post_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "composite",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
    
  # trajs <- plot_indices(inds_comp, 
  #                       add_observed_means = TRUE,
  #                       add_number_routes = TRUE)
  #   
  
  if(mod == "gamye"){
     inds_tmp <- generate_indices(tmp,alternate_n = "n_smooth",
                                    regions = c("Country","bbs_strata"),
                                    regions_index = strata_join,
                                    hpdi = TRUE,
                                    start_year = 1970)
    saveRDS(indices_tmp,paste0("output/inds_comp_smooth_",mod,"_",stratification,".rds"))
    
    indices_tmp <- inds_tmp$indices %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "composite",
             type = "smooth")
    indices_out <- bind_rows(indices_out,
                          indices_tmp)
    
    tt_long <- generate_trends(inds_tmp,
                               min_year = 1970,
                               prob_decrease = c(0,30,50))
    tt_short <- generate_trends(inds_tmp,
                               min_year = 2012,
                               prob_decrease = c(0,30,50))
            tt_pre_90 <- generate_trends(inds_tmp,
                               min_year = 1970,
                               max_year = 1990,
                               prob_decrease = c(0,30,50))
            tt_post_90 <- generate_trends(inds_tmp,
                               min_year = 1990,
                               prob_decrease = c(0,30,50))
    tt_t <- tt_long$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "composite",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
    
    tt_t <- tt_short$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "composite",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
    
        tt_t <- tt_pre_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "composite",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
    
        tt_t <- tt_post_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "composite",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
    
  }

inds_tmp <- generate_indices(tmp,
                         hpdi = TRUE,
                                    start_year = 1970,
                         regions = c("continent","stratum"))

indices_tmp <- inds_tmp$indices %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "full")
    indices_out <- bind_rows(indices_out,
                          indices_tmp)
    
    tt_long <- generate_trends(inds_tmp,
                               min_year = 1970,
                               prob_decrease = c(0,30,50))
    tt_short <- generate_trends(inds_tmp,
                               min_year = 2012,
                               prob_decrease = c(0,30,50))
                tt_pre_90 <- generate_trends(inds_tmp,
                               min_year = 1970,
                               max_year = 1990,
                               prob_decrease = c(0,30,50))
            tt_post_90 <- generate_trends(inds_tmp,
                               min_year = 1990,
                               prob_decrease = c(0,30,50))
            
    tt_t <- tt_long$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
        tt_t <- tt_short$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
        tt_t <- tt_pre_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
        tt_t <- tt_post_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
    
    saveRDS(inds_tmp,paste0("output/indices_",mod,"_",stratification,".rds"))

if(mod == "gamye"){
inds_tmp <- generate_indices(tmp,alternate_n = "n_smooth",
                         hpdi = TRUE,
                                    start_year = 1970,
                         regions = c("continent","stratum"))
saveRDS(inds_tmp,paste0("output/indices_smooth_",mod,"_",stratification,".rds"))

indices_tmp <- inds_tmp$indices %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "smooth")
    indices_out <- bind_rows(indices_out,
                          indices_tmp)
    
    tt_long <- generate_trends(inds_tmp,
                               min_year = 1970,
                               prob_decrease = c(0,30,50))
    tt_short <- generate_trends(inds_tmp,
                               min_year = 2012,
                               prob_decrease = c(0,30,50))
                    tt_pre_90 <- generate_trends(inds_tmp,
                               min_year = 1970,
                               max_year = 1990,
                               prob_decrease = c(0,30,50))
            tt_post_90 <- generate_trends(inds_tmp,
                               min_year = 1990,
                               prob_decrease = c(0,30,50))
            
    tt_t <- tt_long$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
    
    tt_t <- tt_short$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
        tt_t <- tt_pre_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
        tt_t <- tt_post_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
    

}

  
}else{
  inds_tmp <- generate_indices(tmp,
                         hpdi = TRUE,
                                    start_year = 1970,
                         regions = c("continent","country","stratum"))
  indices_tmp <- inds_tmp$indices %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "full")
    indices_out <- bind_rows(indices_out,
                          indices_tmp)
    
    tt_long <- generate_trends(inds_tmp,
                               min_year = 1970,
                               prob_decrease = c(0,30,50))
    tt_short <- generate_trends(inds_tmp,
                               min_year = 2012,
                               prob_decrease = c(0,30,50))
    
                        tt_pre_90 <- generate_trends(inds_tmp,
                               min_year = 1970,
                               max_year = 1990,
                               prob_decrease = c(0,30,50))
            tt_post_90 <- generate_trends(inds_tmp,
                               min_year = 1990,
                               prob_decrease = c(0,30,50))
            
            
    tt_t <- tt_long$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
    
    tt_t <- tt_short$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
        tt_t <- tt_pre_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
        tt_t <- tt_post_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "full")
    trends_out <- bind_rows(trends_out,tt_t)
    
    saveRDS(inds_tmp,paste0("output/indices_",mod,"_",stratification,".rds"))

if(mod == "gamye"){
inds_tmp <- generate_indices(tmp,alternate_n = "n_smooth",
                         hpdi = TRUE,
                                    start_year = 1970,
                         regions = c("continent","country","stratum"))

indices_tmp <- inds_tmp$indices %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "smooth")
    indices_out <- bind_rows(indices_out,
                          indices_tmp)
    
    tt_long <- generate_trends(inds_tmp,
                               min_year = 1970,
                               prob_decrease = c(0,30,50))
    tt_short <- generate_trends(inds_tmp,
                               min_year = 2012,
                               prob_decrease = c(0,30,50))
            tt_pre_90 <- generate_trends(inds_tmp,
                               min_year = 1970,
                               max_year = 1990,
                               prob_decrease = c(0,30,50))
            tt_post_90 <- generate_trends(inds_tmp,
                               min_year = 1990,
                               prob_decrease = c(0,30,50))
            
    tt_t <- tt_long$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
    
    tt_t <- tt_short$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
        tt_t <- tt_pre_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
        tt_t <- tt_post_90$trends %>% 
      mutate(model = mod,
             stratification = stratification,
             summary_type = "simple",
             type = "smooth")
    trends_out <- bind_rows(trends_out,tt_t)
    

saveRDS(inds_tmp,paste0("output/indices_smooth_",mod,"_",stratification,".rds"))

}

}



  }
  }

saveRDS(trends_out,"output/trends_out.rds")
saveRDS(summ_out,"output/summ_out.rds")
saveRDS(indices_out,"output/indices_out.rds")

```

The similarities and differences in trends across the different models and stratifications reflect conclusions and inferences about the species' status about which we should be more or less confident.

The broad-scale status assessments are generally similar across the different models and stratifications.

```{r, echo=FALSE,eval=FALSE}
# compile annual indices


# ind_full_gamye_bbs <- readRDS("output/indices_gamye_bbs_usgs.rds")
# ind_full_gamye_latlong <- readRDS("output/indices_gamye_latlong.rds")
# ind_full_first_diff_bbs <- readRDS("output/indices_first_diff_bbs_usgs.rds")
# ind_full_first_diff_latlong <- readRDS("output/indices_first_diff_latlong.rds")
# ind_full_comp_first_diff_latlong <- readRDS("output/inds_comp_first_diff_latlong.rds")
# ind_full_comp_gamye_latlong <- readRDS("output/inds_comp_gamye_latlong.rds")
# 
# ind_smooth_gamye_bbs <- readRDS("output/indices_smooth_gamye_bbs_usgs.rds")
# ind_smooth_gamye_latlong <- readRDS("output/indices_smooth_gamye_latlong.rds")
# ind_smooth_comp_gamye_latlong <- readRDS("output/inds_comp_smooth_gamye_latlong.rds")
# 

```

## Survey-wide and national trajectories

```{r}

model_names <- data.frame(model = c("first_diff","gamye"),
                          Model = c("First-difference","GAMYE"))
stratification_names <- data.frame(stratification = c("bbs_usgs","latlong"),
                                   Stratification = c("BBS","1-degree cells"))
broad_trajectories <- indices_out %>% 
  filter(region_type %in% c("continent","country","Country"),
         type == "full") %>% 
  mutate(Region = ifelse(region == "continent","Survey-wide",region),
         Region = factor(Region,levels = c("Survey-wide",
                                           "United States of America",
                                           "Canada"))) %>% 
  inner_join(.,model_names,by = "model") %>% 
  inner_join(.,stratification_names,by = "stratification")

trajs <- ggplot(data = broad_trajectories,
                aes(x = year,y = index))+
  geom_ribbon(aes(ymin = index_q_0.05,ymax = index_q_0.95,
                  fill = interaction(Model,Stratification, sep = " : ")), alpha = 0.3)+
  geom_line(aes(colour = interaction(Model,Stratification, sep = " : ")))+
  scale_colour_viridis_d(aesthetics = c("fill","colour"),
                         end = 0.8)+
  scale_y_continuous(trans = "log10")+
  facet_grid(rows = vars(Region),
             scales = "free")+
  guides(fill = guide_legend(title = "Model : Stratification"),
         colour = guide_legend(title = "Model : Stratification"))+
  ylab(expression(Annual~relative~abundance~log[10]))+
  theme_bw()+
  theme(legend.position = "bottom",
        legend.direction = "vertical")

trajs
```

```{r, echo=FALSE}
pdf("figures/trajectories.pdf",
    height = 7.5,
    width = 4)
trajs
dev.off()
```

## Mapping the spatial pattern in trends.

```{r}

strata_trends <- trends_out %>% 
  filter(region_type == "stratum",
         (model == "gamye" & type == "smooth")|
         (model == "first_diff" & type == "full")) %>% 
  mutate(strata_name = region,
         trend_se = width_of_95_percent_credible_interval/(1.96*2),
         span = paste(start_year,end_year,sep = "-"),
         nyears = end_year-start_year,
         span = factor(span,levels = c("1970-2022","1970-1990",
                                       "1990-2022","2012-2022"),
                       ordered = TRUE)) %>% 
  inner_join(.,model_names,by = "model") %>% 
  inner_join(.,stratification_names,by = "stratification")

tcols <- trend_colour_breaks(strata_trends)
pal <- tcols$pal
strata_trends <- strata_trends %>% 
  mutate(trends_map = tcols[["trends_plot_cats"]])

latlong_map <- load_map("latlong") %>% 
  inner_join(.,strata_trends,
            by = "strata_name")


bbs_map <- load_map("bbs_usgs") %>% 
  inner_join(.,strata_trends,
            by = "strata_name")

bbox <- sf::st_bbox(bbs_map)
xlm <- bbox[c("xmin","xmax")]
ylm <- bbox[c("ymin","ymax")]

base_map <- load_map("bbs_usgs")

latlong_trend <- ggplot()+
  geom_sf(data = base_map,
          fill = NA,
          colour = grey(0.7))+
  geom_sf(data = latlong_map,
          aes(fill = trends_map))+
  coord_sf(xlim = xlm,ylim = ylm)+
  scale_fill_manual(values = pal)+
  guides(fill = guide_legend(reverse = TRUE,
                             title = "Trend (%/year)"))+
  theme_bw()+
  facet_grid(cols = vars(Model),
             rows = vars(span))
  
latlong_trend


```

```{r}
latlong_trend_se <- ggplot()+
  geom_sf(data = base_map,
          fill = NA)+
  geom_sf(data = latlong_map,
          aes(fill = trend_se))+
  coord_sf(xlim = xlm,ylim = ylm)+
  scale_fill_viridis_c()+
  guides(fill = guide_colorbar(title = "SE of Trend"))+
  theme_bw()+
  facet_grid(cols = vars(Model),
             rows = vars(span))
  
latlong_trend_se


```

```{r}
pdf("figures/trend_maps_latlong.pdf",
    width = 6,
    height = 10)
latlong_trend
dev.off()

pdf("figures/trend_se_maps_latlong.pdf",
    width = 6,
    height = 10)
latlong_trend_se
dev.off()



```
